\documentclass{article}

\begin{document}
\title{HOPE draft}


\section{Motivation}
We show how to combine the approaches to algebraic effects, together with failures of pattern matching.
We do our development in 3 phases, we show that 

Effect monad ala Kiselyov satisfies the monad laws.
Effect monad ala Kiselyov satisfies the graded monad laws.

\section{Background}

\section{Algebraic effects}



\subsection{Algebraic effects as monads and graded monads}

\subsection{Handling effects}

Handlers of effects can be thought of as a generalization of exception handlers, which provide richer context
and are constrained to respect a certain set of equations.
\subsection{MaybeT and MonadFail}

MaybeT is the monad transformer, i.e. a transformation ${T A : Set} -> Monad T -> Maybe T A -> Monad S$.
$MaybeT m x = m (Maybe x)$
We can think of MaybeT as an extension by an arbitrary, single option failure case. especially in the strucutre of
a Free monad as Monad is Pure and fixed point of a functor with this identity case.


\section{Our results}

We have produced an intermediate construction between the variations of Kiselyov in Agda.
This construction is inbetween the construction the monads Eff and Freer as presented in their work. This is due to our focus on reasoning about monadic laws and
ease of construction within a dependently-typed framework instead of performance.
We further describe how to extend these with failing pattern matching, for simpler description, we present our code in stages and explain how to generically add 
TODO add short explanation of similarities and differences

\section{Effect monad construction}

We review the construction of Effect monad, to illustrate what issues pop up, while trying to make this into a admissible construction in Agda.
In haskell, these constructions can be done easily using the general datatypes, but they included datatype level fixed points in full.
In agda we can take 2 possible views of these constructions.

\section{Graded monad proofs}

We grade monads to allow to combining the effect sets and thus being able to refer to only subcomputations in the programs and proofs.
We do this by describing a reindexing operation, which combines sets and allows us to refer to distinct effects with same names.
We need a way to reindex the eeffect set to embed the left effect set into an ordered monoid.

\section{MaybeT and MaybeFail}
We have to modify handling, we state a different form of initiality, as much as effects are inspired by initiality, we provide a type-theoretic
motivation
for our definition, to provide handling of effectfail, we interpret the results and embed them into Maybe and propagate the underlying failure.
\section{Future directions}
As much we have constructed the monad in agda and verified the various laws. For practical purposes there needs to be more than just a monadic structure.
We want to reason about the local code using the eponymous algebraic laws. We would like to attach equations the handlers should satisfy to the monad.
There are multiple interesting directions to take this in. 
We would also like to use Quotient inductive types to provide a view of operations for reasoning and correct-by-construction definition of handlers,
which abstracts away from the concrete choices when reasoning, yet forces us to prove satisfiability of the algebriac equations by some handler.
To extend the proofs from our current work to the quotient case we need to prove a type-theoretic formulation of algebraic freedom \cite{ncatlab}.
To verify that all the constructions preserve all equations. We note that by theorem 3.1 and 3.2 \cite{ncatlab} this result should follow in Homotopy type theory.
The reason for this being that we need the ambient category to be locally small and complete with homotopy type theory we should be able to get all colimits.

Another direction of future research is to construct an isomorphism between continuations and the fast type-aligned queues of Kiselyov,
thus providing a more efficient implementation of effects without the overheads to reasoning this entails as we focus on semantic constructions instead of executable code, this is not our primary concern.
\end{document}
