\documentclass[10pt]{article}

\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{soul}
\usepackage{sidenotes}
\usepackage{url}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{agda}

% amsthm math environemtsn

\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\newtheorem{lemma}[section]{Lemma}

\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

\theoremstyle{remark}
\newtheorem*{remark}{Remark}

% listing setup

\input{listing-preamble}

\title{EffectFail formalization}
\author{Timotej Tomandl}

% Meta level

\newcommand{\dnoteM}[1]{\textcolor{blue}{\sidenote{\textcolor{blue}{#1}}}}
\newcommand{\dnote}[1]{\textcolor{blue}{#1}}
\newcommand{\tnote}[1]{\textcolor{orange}{#1}}

\newcommand{\ctext}[3][RGB]{%
  \begingroup
  \definecolor{hlcolor}{#1}{#2}\sethlcolor{hlcolor}%
  \hl{#3}%
  \endgroup
}
\newcommand{\highlight}[1]{\ctext[RGB]{255,240,200}{#1}}

\bibliographystyle{plain}


\begin{document}

\maketitle


\begin{abstract}
	When specifying semantics of programming languages, there are many interacting operations be it state, exceptions or environment. In literature these tend to be modelled as monads and combined by using monad transformers.
	In recent years there has been rise in interest in a more modular, as monad transformers are very sensitive to their ordering, called algebraic effects.
	We identify a gap in effect literature in handling partial pattern matching, which is covered by transformers, but incompatible with current approaches to effects.
	We present our formalization of algebraic effects in Agda combined with the ability to have partial patterns.
\end{abstract}

\section{Introduction}

% introduce effect semantics via handlers vs. monads
Functional programming presents an opportunity to reconsider the role
of side effects in programs. There have been various techniques
proposed for integrating effects smoothly into functional code.
The predominant approaches are \emph{algebraic effects} and the
\emph{monadic} approach. One of the motivations behind algebraic
effects is the desire to allow the management of side effects
to be \emph{extensible}.

Monad transformers start with an encoding and only afterwards attempt to provide the necessary operations. Effects do the reverse we firt think in terms of operation and equations, which the
y should satisfy and only provide implementation afterwards. An advantage of it is that the interpretation does not change structure of already written code, while this tends to happen when adding new operations using transformers. \cite{exteff}

Algebraic effects provide a modular and lower overhead construction to
flexibly add new operations to a monad, compared to a monad
transformer approach.
%
In monad transformer frameworks we need a new type class instance for every
combination of effects and this makes the implementation less
flexible.

Algebraic effects
provide ways to write programs causing effects which are treated
abstractly initially but which are given concrete implementations
afterwards using \emph{handlers} (akin to raising an exception and
later handling it).
This generalizes to general effects: effect handlers provide a way to
not only handle general exceptions, but to also remember some state between handling different operations a handler is responsible for.
For example, this allows implementing stateful computation, which replies to a get effect by returning the last put effects input.
Algebraic effect handlers conform to a standard interface, which we call the 'effect API', comprising the use of an effectful operation (called a 'send', akin to throw for exceptions) and a way to interpret, or 'handle', that operation (akin to catch for exceptions). We define later in more precise terms the form of the effect API, which are given in a generic way for any operation.
Thus an effect API provides a way to `send' effects and `handle' them, where `send' corresponds to throwing and `handle' to catching an exception. It also provides a way to extract the final value once we have handled all effects.

One of our larger aims is to explore how to emulate modular monadic semantics \cite{Schrijvers}, which provides composable way to combine language feature with monadic semantics approach, which allows expressing semantics in terms of operations provided by transformers. We hope this gives a good start towards managing complexities, which arise with pattern matching and algebraic effects.

\paragraph{Problem statement}
% jump from context/techniques to the problem
Whilst algebraic effects are growing in maturity as a technique, there
still exists some pain points with regards their practical use when
compared with monads and monad transformers. In this paper, we
study the problem of \emph{failure}, especially with regards
managing pattern matching failures. In Haskell, a well-used technique
is the \texttt{MonadFail} pattern, which captures pattern match failure
as an effect for a particular monad to reduce boilerplate.

The handlers also inform our encodings, when we send an effect handlers must have convenient way to resume computation by replying to it and they must be able to easily identify whether a given effect is meant for them.

However, the central approach of algebraic effects uses the
free monad construction,
which does not admit the \texttt{MonadFail} pattern immediately.
The free monad construction is used for imposing exacty the minmal amount of structure necessary for an object to be a monad, all other structures will impose extra laws, while free monad provides only the monad laws.

We want not only to use the Effect API for monads
but also give a semantics of pattern matching that may fail.
As such we set out to find a monad supporting both failure and the Effect API.

% OLD TEXT
% Standard formulations of effect monads can not support the
% interface of MonadFail\dnoteM{First mention of this-- needs
% introducing first} as they are, because they use a Free Monad
% construction, which gives the simplest construction for which Monad
% laws hold.


% grading

% summarise the reuslts
We make the following contributions:
%
\begin{enumerate}
\item We refine a standard free monad construction to a form that is
  suitable for formalisation in a proof assistant---we use Agda. We
  prove this is still a monad.

\item We show that applying the standard \texttt{MaybeT} monad
  transformer to our free monad construction yields a lawful
(i.e., well behaved) \texttt{MonadFail} instance, allowing
it to capture pattern matching failure.

\item Furthermore, we show that this construction also
yields a graded monad, giving a typed representation of
the effects yielded by a computation.

\item We provide a partial implementation and verification of these ideas in Agda \ref{repo}

% ...
\end{enumerate}
%
We name this approach \textit{EffectFail} since it combines the idea
of a free monad for capturing algebraic effects with the
\texttt{MonadFail} pattern of Haskell.

In Section 2, we discuss the background of our work
In Section 3, we discuss our construction of effect monads and prove that it is a monad and a graded monad.
In Section 3, we provide the implementation of the Effect API for this monad.
In Section 4, we construct the \emph{EffectFail} interface, showing
that it satisfies the \texttt{MonadFail} laws.
In Section 5, we discuss future directions and applications we have in mind for this formalization.


% context of this use case
%

\subsection{Motivating example}

We also want to provide a way to match on
patterns and fail in a default way, if they are not
satisfied.

\paragraph{Wider context}

Patterns play a crucial role in semantics definition of intermediate
languages, such as CakeML compiler intermediate representations.
After some point in the compiler we forget the information about typing
and make certain operations just fail if provided with wrong inputs.
One example from the CakeML compiler being the following,
%
\begin{lstlisting}[language=Caml,basicstyle=\ttfamily\small]
      | (El,[RefPtr ptr; Number i]) =>
        (case FLOOKUP s.refs ptr of
         | SOME (ValueArray xs) =>
            (if 0 <= i /\ i < & (LENGTH xs)
             then Rval (EL (Num i) xs, s)
             else Error)
         | _ => Error)
\end{lstlisting}
%
Here we lookup a value of pointer ptr and require it contains a specific kind of an array, if any other kind of an array is being pointed to we fail in a default way.
Same happens, when we get an out-of-bounds error, in the if case.
We would like to rewrite this using Haskell-style do notation as follows,
%
\begin{lstlisting}[language=Caml,basicstyle=\ttfamily\small]
	| (El,[RefPtr ptr; Number i]) => do
  	  (ValueArray xs) <- FLOOKUP s.refs ptr
	  cond (0 <= i /\ i < & (LENGTH xs))
	  return (Rval (EL (Num i) xs , s))
\end{lstlisting}
%

We can further refine this using state monad,
%
\begin{lstlisting}[language=Caml,basicstyle=\ttfamily\small]
      | (El,[RefPtr ptr; Number i]) => do
      	  (ValueArray xs) <- FLOOKUP ptr
	  cod (0 <= i /\ i < & (LENGTH xs))
	  return (Rval (EL (Num i) xs))
\end{lstlisting}
%
This hides the state passing of s from users view. But there are other operations, not everything is looked up through state, some operations might lookup in a current environment or throw an exception. As such we would like to combine differet operations in a single monad.
\section{Background} % Free monads and MonadFail}
%
\subsection{Monads, MonadFail and failing pattern matching}
%
Haskell provides a syntactic sugar for composing monadic computations,
called the ``do'' notation~\cite{jones2003haskell}, that is similar to Moggi's monadic
metalanguage~\cite{moggi1991notions}. The basic form is akin to a kind
of let-binding:
%
\begin{haskell}
     do
	x <- action1
	action2
\end{haskell}
%
which desugars to the following Haskell code:
%
\begin{haskell}
    action1 >>= (\ x -> action2)
\end{haskell}
%
This requires, that terms \haskin{action1} and \haskin{action2} both return
a value of a shared monadic type, e.g. \haskin{action1 :: m a} and
\haskin{action2 :: m b} for some monadic type \haskin{m}. The
\haskin{>>=} operation is pronouced \emph{bind}, providing the key
notion of monadic sequential composition: \haskin{(>>=) :: m a -> (a
  -> m b) -> m b}. The notation \haskin{x <- ...} is often also then
referred to as a ``bind''.

The pattern matching bind,
%
\begin{haskell}
     do
	pat <- action1
	action2
\end{haskell}
%
where pat is not a variable but a pattern
with possible variables included is an extension of this
idea,
desugars into,
%
\begin{haskell}
    action1 >>= (\ x -> case x of
	             pat -> action2
		     _ -> fail)
\end{haskell}
%

Here x is selected as not to conflict with any user-defined variables. Not only does this desugaring require monadic bind, but also a new operation fail.
This is captured in the Haskell type class MonadFail, which adds this
operation.

The type of \haskin{fail} in Haskell is $String \rightarrow m\:a$ for some monad
$m$. We are interested in formulation not passing error
messages, as these are used in Haskell to identify the reason, why the pattern matched fail and potetially print out errors. These are ignored by most interesting instances in the I/O free setting.


In haskell it is assumed due to the expected semantics of pattern matching that fail satisfies the following law, for all x,
%
\begin{haskell}
	fail >>= x = fail
\end{haskell}
%
i.e., once a failure has occured, as given by \haskin{mfail}, then the
entire computation fails--subsequent sequentially composed computing
(e.g., \haskin{f} above) are ignored.
\dnote{TODO:
  integrate references and context
  \cite{gibbons2011just},\cite{malakhovski2018exceptionally},
\cite{monadfailproposal} SCitations to its use in the literature and discuss a bit what
  has been said about it there.}

Thus, from the view of monads as monoids (in the category of
endofunctors), \haskin{mfail} identifies an element of the monoid
which is \emph{cancellative} or \emph{annihilating} with respect
to sequential composition; local failure becomes
global failure.

\subsection{Free monads and algebraic effects}
To provide algebraic effects, the normally used construction is that of a free monad.
The reason for this is that it provides the simplest monad possible, for a given functor.
This allows for imposing only laws of monads and nothing more, this is important as we do not want to impose any laws, which do not come from the effects. This is further followed by lifting the restriction that we need a functor, by using a construction, which gives functor for any type family, this allows holding exactly one effect. This can be then refined to hold any effect out of certain collection.

\subsubsection{Standard free monad construction can not provide mfail}

If we attempt to give a \texttt{MonadFail} instance for the free monad we get
stuck on the conceptual level already: the free monad, being the monad for
which exactly the monad equations hold and nothing more on one hand
and \texttt{MonadFail} on the other hand, which adds an extra equation.
Here we focus on forming a reasonable API supporting both effect handling and failure.
The simplest construction supporting the new law generically is
\texttt{MaybeT}, which we focus on in the next section. We conjecture
that our construction is a free \texttt{MonadFail} (see
Section~\ref{sec:future-work})

\subsection{Effect API}

Effect API consists of 2 pieces handling and sending an effect, we run
into an issue with Handling though, when attempting to handle an
effect we have to do Set comparison, which is not available in Agda.
API for effects consists of the following.

\begin{agda}
send : eff a -> Effect effs a -> Effect (insert eff effs) a
\end{agda}

this sends a new effect inside a monad

\begin{agda}
handle : Effect effs a -> (a -> Effect effs2 b) -> (forall eff. eff ->
Effect effs2 b) -> Effect effs2 b
\end{agda}

interprets one effect set inside another effect set

\begin{agda}
escape : Effect [] a -> a
\end{agda}

this extracts a final value once all effects have been handled.
Send let's us write computation causing abstract effects,
handle gives concrete interpretation to this abstraction and escape finishes once we have
concretized interpreted all possible effects by handling.
The most interesting part of this is handle, it let's us reinterpter
effectful computation in terms of other effectful computations and gradually give a concrete implementation.

\section{A free monad construction for algebraic effects}
\label{sec:effect-monad}

\subsection{Standard construction}

Standard Haskell implementation of algebraic effects,  following Kiselyov \cite{Kiselyov},
tend to be conceived in two steps: constructing a non-strictly positive
\haskin{Free} monad and proving that it is a monad for any functor,
then instantiating it with a Left Kan extension, which gives us the functor for any type family.For an implementation, these two steps
tend to be fused together and instead of requiring the left Kan
extension argument is instantiated with the type.

\begin{haskell}
	data Free (F : * -> *) A where
		Pure : A -> Free F A
		Impure : F (Free F A) -> Free F A
\end{haskell}
%

The above Free monad gives us a monad for any functor F, being the simplest such construction, satisfying only the laws imposed by Functor instance of F and the monadic laws.

Then we instatiate F with Left kan extension of G, which gives us functor for any G, thus getting a monad for any G.

\begin{haskell}
	data Lan (G : * -> *) A where
		FMap : (B -> A) -> G B -> Lan G A
\end{haskell}
%

When mapping this construction just remembers the arguments and accumulates functions, which were mapped.

We would like to do these conceptually separately for formalization, but we run into a snag, Free monad is not strictly positive and as such can not be consistently implemeted in Agda. The instantiation of Free with the left Kan extension is strictly positive as such we just use the instantiation directly.

While previous approaches \cite{kiselyov} aim mostly for efficiency of implementation we aim for ease of proofs. As such we take intermediate point in the design space of this paper.

We just require that
it is instantiated with a type, which occurs within a type-level list
at a given index stored within the datatype. Being in type-level list
is then enforced through type class constraints.

The construction works by looking for inspiration to category theory and taking the free structure for monadic equations, which gives a generic construction which satisfies monadic equations for every functor. This gives a structure, which satisfies the monad laws, but does not impose any extra laws by itself, this gives us a way to use monadic notation and reasoning and imposes only the expected restrictions.
It then looks to elevate the requirement of functoriality by taking a generic construction, which returns a functor for any type family.

We show our construction, step-by-step. Our construction is adapted from the work of Kyseliov \cite{freerm} and traces back to using free monads to represent uninterpreted effects, the idea is that Free monad imposes the minimum amount of constraints, while still being a monad, in other words the only laws that should hold for it are the monad laws, with rest of the operations being unconstrained.

We start with the construction of a Free monad and Left Kan extension as above and instantiate Free with the Left Kan extesion, this yields the following.

\begin{haskell}
	data EffP (F : * -> *) A where
		Pure : A -> EffP F A
		Impure : (B -> EffP F A) -> F B -> EffP F A
\end{haskell}
%

We then then refine it to hold any \haskin{E} from within a type-level list \haskin{F}.
\begin{haskell}
      data Eff (F : [* -> *] A where
              Pure : A -> Eff F A
              Impure : (Member F E) => (B -> Eff F A) -> E B -> Eff F A
\end{haskell}
%

Being in type-level list
is then enforced through type class constraints.


\subsection{Our construction in Agda}

We have formalized our constructions in Agda \cite{githubefffail}.
We are working in dependently typed setting without type classes and
as such we need to do multiple tweaks. Without some notion of
positive and negative occurences,
we have to do the instantiation above in a single step.
Also Agda does not provide powerful enough type class resolution
mechanism and this would be unsatisfyingly metatheoretic. We construct
our effect monad by tweaking the Haskell construction slightly and
instead of providing an index to type-level list, we take a proof of
contained effect's membership in this list. As proofs of membership in
Agda are proof-relevant this at the same time acts as an index.
\begin{agda}
	data Effect (E : List (Set -> Set)) (A : Set) : Set1 where
	  Pure : A -> Effect E A
	  Impure : {F : Set} F X
	     -> (F in E)
	     -> (X -> Effect E A)
	     -> Effect E A
\end{agda}
Our construction changes the typeclass resolution constraint of Effect in Haskell to a proof that F occurs in the effect list.
We would like to further refine this to provide names to every effect in the list
and thus being able to refer to each effect by a unique name, instead
of by its type.

\subsection{Algebraic effects as graded monads}

The concept of \emph{graded monads} generalises the notion of monads
from a single endofunctor (e.g., used to model a class of effects) to
an indexed family of endofunctors, essentially stratifying a model of
effects into different sub-models distinguished by the
index~\cite{DBLP:conf/popl/Katsumata14,DBLP:journals/corr/OrchardPM14}. The
indices have the structure of a monoid or pre-ordered monoid, which is
leveraged in the graded monad operations which are analogues of the
usual monad operations (and axioms).

Having graded monad laws is useful for combining computations producing different effects. While normal monadic bind only supports combining computations with the same effect set, now we can write our computations independently in the smallest effect set and combine them to produce a new computation with a larger effect set. It consists of operations with similar signatures to bind and pure of our effect monad, but now annotated differently. Graded pure produces only effects with empty set. Graded bind on the other hand produces an effect, which combines the operations by list concatentation. This means our subcomputations are not forced to know the whole set of operations in advance.

In future we would like to quotient our list of effects to remove duplicated names and thus obtaining graded monad, which is much closer to the instance resolution notion handling-style.


\begin{definition}[Graded monad]
	Graded monad has similar structure as a monad except it's elements are annotated by elements from a monoid. Monoid being simple a structure on a set S,
	consisting of an associative binary operations $m\: :\:S\:\times\:S\:\rightarrow\:S$, and a distinguished element $E$ in $S$ which acts as both left and right unit.
	Graded return operation is annotated by the monoid's unit and graded bind uses the monoids associative operation to combine the annotations of subcomputations.
\end{definition}

\begin{proposition}[\texttt{Effect} is a graded monad]
\end{proposition}

The proof has been mechanised in Agda.
This gives a compositional notion of binding and purity.

\section{Maybe transformer gives an instance of MonadFail}

Now we focus on the next piece and that's giving the MonadFail instance to our Effect monad,
which justifies pattern matching failures
A standard approach of adding a notion of failure to an existing
monad is via a \emph{monad transformer}, typically
\haskin{MaybeT}~\cite{transformerpub}:
%
\begin{haskell}
newtype MaybeT m a = MaybeT { runMaybeT :: m (Maybe a) }
\end{haskell}
%
We prove that this transformer is a monad for any monad \haskin{m}.
We also prove that it is a lawful instance of \haskin{MonadFail} for
any monad \haskin{m}.
The first construction was introduced together with monad transformers
\cite{transformerpub}, the MonadFail instance in various Haskell
implementations \cite{mtl}. We give these constructions in Agda and
verify the laws of \haskin{Monad} and \haskin{MonadFail} given that m is a \haskin{Monad}.

\subsection{Effect API through Maybe transformer}

We lift the Effect API through Maybe transformer by identifying
sending with embedding Just values inside Effect, this has the desired
effect of being able to handle them properly. We then handle in the
same fashion. Once all effects are handled and the list of
possibilities is empty we have to extract the pure value, to do this
we note that impure case requires proof of membership inside the list
and this is not possible for the empty list, thus we extract the optional result from
Pure.

\section{Future directions}
\label{sec:future-work}

As much as we have implemented an effect API we would like to also reason about it.
One possible refinement is for handlers to not only take the instances of a given effect, but also quotient them to enforce that algebraic laws hold \cite{algtheorieshit}. This would force us to provide correctness proofs whenever mapping out of a particular effect and ensure that laws used to reason locally are enforced by handlers.
We would like to explore how to use category theoretic constructions relating initial algebras to quotient types, to encode the relationship between free algebras and their handlers.
This would involve proving that our construction is the free algebra with respect to the algebraic signature of MonadFail laws and then relating the quotiented variants to interpretations via initial constructions.

\bibliography{main}

\end{document}
